---
title: "Inspect FishGlob — NS-IBTS"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This notebook loads **`NS-IBTS_clean.RData`** from the FishGlob cleaned outputs (North Sea International Bottom Trawl Survey) and summarizes what is inside.

## Locate and load data

Paths work when the notebook lives in **`north_sea_eeos`**, or when it is in a subfolder such as **`NS_EEOS`** (parent is searched automatically).

```{r load}
find_ns_ibts <- function() {
  candidates <- c(
    file.path("FishGlob_data", "outputs", "Cleaned_data", "NS-IBTS_clean.RData"),
    file.path("..", "FishGlob_data", "outputs", "Cleaned_data", "NS-IBTS_clean.RData")
  )
  for (p in candidates) {
    if (file.exists(p)) return(normalizePath(p))
  }
  stop(
    "NS-IBTS_clean.RData not found. Expected under north_sea_eeos/FishGlob_data/outputs/Cleaned_data/."
  )
}

path_rdata <- find_ns_ibts()
cat("Loaded from:\n", path_rdata, "\n\n")

load(path_rdata)
# Objects: `data` (main table), `readme` (metadata)
```

## Dataset readme (`readme`)

```{r readme}
readme
```

## Table dimensions and column names

```{r dims}
cat("Rows:", nrow(data), "  Columns:", ncol(data), "\n")
names(data)
```

## Quick structure

```{r glimpse, results='asis'}
if (requireNamespace("pillar", quietly = TRUE) && requireNamespace("tibble", quietly = TRUE)) {
  print(tibble::as_tibble(utils::head(data, 5)))
} else {
  knitr::kable(utils::head(data, 5), row.names = FALSE)
}
```

Use **`dplyr::glimpse(data)`** in the console for a full variable-wise summary if you use the tidyverse.

## Time and sampling coverage

```{r coverage}
ys <- range(data$year, na.rm = TRUE)
n_haul <- length(unique(data$haul_id))
n_taxa <- length(unique(data$accepted_name[!is.na(data$accepted_name)]))
cat("Year range:", ys[1], "–", ys[2], "\n")
cat("Unique haul_id:", n_haul, "\n")
cat("Unique accepted species names (non-NA):", n_taxa, "\n")
```

## Records per year

```{r per-year}
yr <- table(data$year)
yr_df <- data.frame(year = as.integer(names(yr)), n_records = as.integer(yr))
yr_df <- yr_df[order(yr_df$year), ]
knitr::kable(utils::head(yr_df, 15), caption = "First 15 years (rows = species × haul records)")
knitr::kable(utils::tail(yr_df, 15), caption = "Last 15 years")
```

## Top species by total catch weight (`wgt`)

```{r top-species}
agg <- aggregate(wgt ~ accepted_name, data = data, FUN = sum, na.rm = TRUE)
names(agg)[2] <- "total_wgt"
agg <- agg[order(agg$total_wgt, decreasing = TRUE), ]
rownames(agg) <- NULL
knitr::kable(utils::head(agg, 20), caption = "Top 20 species by summed wgt")
```

## Spatial snapshot (first and last years with coordinates)

```{r map-table}
xy <- data[!is.na(data$latitude) & !is.na(data$longitude), c("year", "latitude", "longitude", "haul_id")]
xy <- unique(xy)
if (nrow(xy) > 0) {
  ylo <- min(xy$year)
  yhi <- max(xy$year)
  s1 <- xy[xy$year == ylo, ][1:min(5, sum(xy$year == ylo)), , drop = FALSE]
  shi <- xy[xy$year == yhi, ][1:min(5, sum(xy$year == yhi)), , drop = FALSE]
  knitr::kable(s1, caption = paste("Example hauls (first year in data:", ylo, ")"))
  knitr::kable(shi, caption = paste("Example hauls (last year in data:", yhi, ")"))
}
```

To plot points in R, use e.g. **`plot(data$longitude, data$latitude, pch = 16, cex = 0.2, asp = 1)`** or **`ggplot2::geom_point`** with alpha for density.

---

## Quarter overview

How does the data split between quarters overall?

```{r quarter-overview}
library(dplyr)
library(knitr)

data %>%
  count(quarter) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  kable(caption = "Records by quarter (species × haul rows)")
```

---

## Q1 vs Q3 — annual haul counts

Number of unique hauls per year, split by quarter. This is the primary comparator for temporal consistency.

```{r q1q3-hauls-year}
hauls_by_quarter_year <- data %>%
  filter(quarter %in% c(1, 3)) %>%
  distinct(quarter, year, haul_id) %>%
  count(quarter, year, name = "n_hauls")

# Wide format for easy comparison
hauls_wide <- hauls_by_quarter_year %>%
  tidyr::pivot_wider(names_from = quarter, values_from = n_hauls,
                     names_prefix = "Q") %>%
  arrange(year)

kable(hauls_wide, caption = "Unique hauls per year: Q1 vs Q3")
```

---

## Q1 vs Q3 — species richness per year

Number of unique species observed per year, by quarter.

```{r q1q3-species-year}
data %>%
  filter(quarter %in% c(1, 3), !is.na(accepted_name)) %>%
  group_by(quarter, year) %>%
  summarise(n_species = n_distinct(accepted_name), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = quarter, values_from = n_species,
                     names_prefix = "Q") %>%
  arrange(year) %>%
  kable(caption = "Unique species per year: Q1 vs Q3")
```

---

## Q1 vs Q3 — spatial coverage per year

Number of unique ICES rectangles (stat_rec) sampled per year.

```{r q1q3-rects-year}
data %>%
  filter(quarter %in% c(1, 3), !is.na(stat_rec)) %>%
  distinct(quarter, year, stat_rec) %>%
  count(quarter, year, name = "n_rects") %>%
  tidyr::pivot_wider(names_from = quarter, values_from = n_rects,
                     names_prefix = "Q") %>%
  arrange(year) %>%
  kable(caption = "Unique ICES rectangles sampled per year: Q1 vs Q3")
```

---

## Q1 vs Q3 — summary statistics

Single-row summary comparing the two quarters across the full time series (year >= 1983).

```{r q1q3-summary}
data %>%
  filter(quarter %in% c(1, 3), year >= 1983, !is.na(stat_rec), !is.na(accepted_name)) %>%
  group_by(quarter, year) %>%
  summarise(
    n_hauls   = n_distinct(haul_id),
    n_rects   = n_distinct(stat_rec),
    n_species = n_distinct(accepted_name),
    .groups   = "drop"
  ) %>%
  group_by(quarter) %>%
  summarise(
    years_with_data    = n(),
    first_year         = min(year),
    last_year          = max(year),
    median_hauls_yr    = median(n_hauls),
    sd_hauls_yr        = round(sd(n_hauls), 1),
    median_rects_yr    = median(n_rects),
    median_species_yr  = median(n_species),
    .groups = "drop"
  ) %>%
  kable(caption = "Summary across years >= 1983: Q1 vs Q3")
```

---

## Haul density per rectangle per year (Q1, year >= 1983)

Distribution of hauls-per-cell-per-year — the key statistic for deciding the minimum density threshold.

```{r haul-density-q1}
density_q1 <- data %>%
  filter(quarter == 1, year >= 1983, !is.na(stat_rec)) %>%
  distinct(year, stat_rec, haul_id) %>%
  count(year, stat_rec, name = "n_hauls")

cat("=== Q1: hauls per rectangle per year (year >= 1983) ===\n")
summary(density_q1$n_hauls)

cat("\nDistribution of haul counts:\n")
density_q1 %>%
  count(n_hauls, name = "n_cell_years") %>%
  mutate(cumulative_pct = round(100 * cumsum(n_cell_years) / sum(n_cell_years), 1)) %>%
  kable(caption = "How often does each haul count occur across all cell-years?")
```

---

## Haul density per rectangle per year (Q3, year >= 1983)

Same table for Q3 — compare directly with Q1 above.

```{r haul-density-q3}
density_q3 <- data %>%
  filter(quarter == 3, year >= 1983, !is.na(stat_rec)) %>%
  distinct(year, stat_rec, haul_id) %>%
  count(year, stat_rec, name = "n_hauls")

cat("=== Q3: hauls per rectangle per year (year >= 1983) ===\n")
summary(density_q3$n_hauls)

cat("\nDistribution of haul counts:\n")
density_q3 %>%
  count(n_hauls, name = "n_cell_years") %>%
  mutate(cumulative_pct = round(100 * cumsum(n_cell_years) / sum(n_cell_years), 1)) %>%
  kable(caption = "How often does each haul count occur across all cell-years?")
```

---

## Cells passing density thresholds (Q1 vs Q3)

At various minimum haul thresholds, how many cell-years survive? Informs threshold choice.

```{r density-thresholds}
thresholds <- c(1, 2, 3, 5, 8, 10)

thresh_compare <- purrr::map_dfr(thresholds, function(t) {
  tibble(
    min_hauls = t,
    Q1_cell_years  = sum(density_q1$n_hauls >= t),
    Q1_pct         = round(100 * mean(density_q1$n_hauls >= t), 1),
    Q3_cell_years  = sum(density_q3$n_hauls >= t),
    Q3_pct         = round(100 * mean(density_q3$n_hauls >= t), 1)
  )
})

kable(thresh_compare,
      caption = "Cell-years surviving each minimum haul threshold: Q1 vs Q3")
```

---

## Temporal consistency per rectangle (Q1, year >= 1983)

For each rectangle: how many years does it appear in Q1 with >= 3 hauls? Identifies the stable core panel.

```{r rect-temporal-q1}
rect_years_q1 <- density_q1 %>%
  filter(n_hauls >= 3) %>%
  count(stat_rec, name = "n_years_ge3hauls") %>%
  arrange(desc(n_years_ge3hauls))

cat("Q1 rectangles by number of years with >= 3 hauls:\n")
rect_years_q1 %>%
  count(n_years_ge3hauls, name = "n_rectangles") %>%
  arrange(desc(n_years_ge3hauls)) %>%
  mutate(cumulative_rects = cumsum(n_rectangles)) %>%
  kable(caption = "How many rectangles have N or more years with >= 3 hauls? (Q1)")

cat("\nRectangles with >= 20 years of Q1 data (>= 3 hauls/yr):", 
    sum(rect_years_q1$n_years_ge3hauls >= 20), "\n")
cat("Rectangles with >= 30 years of Q1 data (>= 3 hauls/yr):", 
    sum(rect_years_q1$n_years_ge3hauls >= 30), "\n")
```

---

## Temporal consistency per rectangle (Q3, year >= 1983)

Same for Q3.

```{r rect-temporal-q3}
rect_years_q3 <- density_q3 %>%
  filter(n_hauls >= 3) %>%
  count(stat_rec, name = "n_years_ge3hauls") %>%
  arrange(desc(n_years_ge3hauls))

cat("Q3 rectangles by number of years with >= 3 hauls:\n")
rect_years_q3 %>%
  count(n_years_ge3hauls, name = "n_rectangles") %>%
  arrange(desc(n_years_ge3hauls)) %>%
  mutate(cumulative_rects = cumsum(n_rectangles)) %>%
  kable(caption = "How many rectangles have N or more years with >= 3 hauls? (Q3)")

cat("\nRectangles with >= 20 years of Q3 data (>= 3 hauls/yr):", 
    sum(rect_years_q3$n_years_ge3hauls >= 20), "\n")
cat("Rectangles with >= 30 years of Q3 data (>= 3 hauls/yr):", 
    sum(rect_years_q3$n_years_ge3hauls >= 30), "\n")
```

---

## Spatial coverage maps (Q1, year >= 1983)

Four maps showing the spatial structure of the panel under the ≥2 hauls sub-sampling rule.
Requires `ggplot2`, `sf`, and `rnaturalearth`.

```{r spatial-maps, fig.width=10, fig.height=9}
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# ── Step 1: derive rectangle centroids from haul coordinates in the data ──────
# Average haul lat/lon within each stat_rec — accurate to within a fraction
# of a degree since all hauls in a rectangle share the same ~0.5° x 1° cell.
coords <- data %>%
  filter(quarter == 1, year >= 1983,
         !is.na(stat_rec), !is.na(latitude), !is.na(longitude)) %>%
  group_by(stat_rec) %>%
  summarise(
    lon = mean(longitude, na.rm = TRUE),
    lat = mean(latitude,  na.rm = TRUE),
    .groups = "drop"
  )

# ── Step 2: Q1 cell-years with >= 2 hauls ─────────────────────────────────────
panel_rects <- data %>%
  filter(quarter == 1, year >= 1983, !is.na(stat_rec)) %>%
  distinct(year, stat_rec, haul_id) %>%
  count(year, stat_rec, name = "n_hauls") %>%
  filter(n_hauls >= 2)

# ── Step 3: rectangle-level summaries, joined to coords ───────────────────────
rect_summary <- panel_rects %>%
  count(stat_rec, name = "n_years") %>%
  mutate(
    temporal_class = case_when(
      n_years >= 30 ~ "≥30 yrs",
      n_years >= 20 ~ "20–29 yrs",
      n_years >= 10 ~ "10–19 yrs",
      TRUE          ~ "<10 yrs"
    ),
    temporal_class = factor(temporal_class,
                            levels = c("≥30 yrs", "20–29 yrs", "10–19 yrs", "<10 yrs"))
  )

rect_density <- panel_rects %>%
  group_by(stat_rec) %>%
  summarise(mean_hauls = mean(n_hauls), .groups = "drop")

rect_plot <- rect_summary %>%
  left_join(rect_density, by = "stat_rec") %>%
  left_join(coords, by = "stat_rec")

# ── Map 1: Years in panel (temporal depth) ────────────────────────────────────
p1 <- ggplot() +
  geom_sf(data = coast, fill = "grey88", colour = "grey60", linewidth = 0.3) +
  geom_tile(data = rect_plot,
            aes(x = lon, y = lat, fill = temporal_class),
            width = 1, height = 0.5, alpha = 0.85) +
  scale_fill_manual(
    values = c("≥30 yrs" = "#0C447C", "20–29 yrs" = "#378ADD",
               "10–19 yrs" = "#85B7EB", "<10 yrs" = "#E6F1FB"),
    name = "Years in panel"
  ) +
  coord_sf(xlim = ns_xlim, ylim = ns_ylim, expand = FALSE) +
  labs(title = "Temporal depth per rectangle",
       subtitle = "Q1, year ≥ 1983, ≥ 2 hauls/cell/year") +
  map_theme

# ── Map 2: Continuous years in panel ─────────────────────────────────────────
p2 <- ggplot() +
  geom_sf(data = coast, fill = "grey88", colour = "grey60", linewidth = 0.3) +
  geom_tile(data = rect_plot,
            aes(x = lon, y = lat, fill = n_years),
            width = 1, height = 0.5) +
  scale_fill_gradient(low = "#E6F1FB", high = "#042C53",
                      name = "Years\nin panel") +
  coord_sf(xlim = ns_xlim, ylim = ns_ylim, expand = FALSE) +
  labs(title = "Years in panel (continuous)",
       subtitle = "Q1, year ≥ 1983, ≥ 2 hauls/cell/year") +
  map_theme

# ── Map 3: Mean hauls per cell-year ──────────────────────────────────────────
p3 <- ggplot() +
  geom_sf(data = coast, fill = "grey88", colour = "grey60", linewidth = 0.3) +
  geom_tile(data = rect_plot,
            aes(x = lon, y = lat, fill = mean_hauls),
            width = 1, height = 0.5) +
  scale_fill_gradient(low = "#E1F5EE", high = "#04342C",
                      name = "Mean\nhauls/yr") +
  coord_sf(xlim = ns_xlim, ylim = ns_ylim, expand = FALSE) +
  labs(title = "Mean hauls per cell-year",
       subtitle = "Q1, year ≥ 1983, ≥ 2 hauls/cell/year") +
  map_theme

# ── Map 4: First year each rectangle enters the panel ────────────────────────
rect_first_year <- panel_rects %>%
  group_by(stat_rec) %>%
  summarise(first_year = min(year), .groups = "drop") %>%
  left_join(coords, by = "stat_rec")

p4 <- ggplot() +
  geom_sf(data = coast, fill = "grey88", colour = "grey60", linewidth = 0.3) +
  geom_tile(data = rect_first_year,
            aes(x = lon, y = lat, fill = first_year),
            width = 1, height = 0.5) +
  scale_fill_gradient(low = "#042C53", high = "#85B7EB",
                      name = "First\nyear") +
  coord_sf(xlim = ns_xlim, ylim = ns_ylim, expand = FALSE) +
  labs(title = "First year rectangle enters panel",
       subtitle = "Q1, year ≥ 1983, ≥ 2 hauls/cell/year") +
  map_theme

# ── Combine into 2x2 grid ─────────────────────────────────────────────────────
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title    = "NS-IBTS Q1 panel spatial structure",
    subtitle = "182 rectangles · 5,643 cell-years · median 34 years per rectangle",
    theme    = theme(
      plot.title    = element_text(size = 13, face = "bold"),
      plot.subtitle = element_text(size = 10, colour = "grey40")
    )
  )
```

---

## Spatial coverage — annual snapshots

How does rectangle coverage shift across three decades? Useful for spotting
whether peripheral rectangles drop in/out systematically.

```{r spatial-snapshots, fig.width=11, fig.height=7}
# Coverage in selected years
snapshot_years <- c(1983, 1990, 1995, 2000, 2005, 2010, 2015, 2020)

snap_data <- panel_rects %>%
  filter(year %in% snapshot_years) %>%
  left_join(coords, by = "stat_rec")

ggplot() +
  geom_sf(data = coast, fill = "grey88", colour = "grey60", linewidth = 0.2) +
  geom_tile(data = snap_data,
            aes(x = lon, y = lat),
            width = 1, height = 0.5,
            fill = "#185FA5", alpha = 0.75) +
  coord_sf(xlim = ns_xlim, ylim = ns_ylim, expand = FALSE) +
  facet_wrap(~ year, nrow = 2) +
  labs(
    title    = "Rectangle coverage in selected years (Q1, ≥ 2 hauls)",
    subtitle = "Blue = rectangle present in panel that year"
  ) +
  map_theme +
  theme(axis.text = element_text(size = 7))
```

---

## Column availability — length and weight (Q1, year >= 1983)

Confirms whether length and weight data are present and how well they are covered,
which determines the feasibility of individual-level E calculation and B reliability.

```{r column-availability}
q1_83 <- data %>%
  filter(quarter == 1, year >= 1983)

cat("=== Total rows (species x haul records) ===\n")
cat("Q1, year >= 1983:", nrow(q1_83), "\n\n")

# ── Which columns exist? ───────────────────────────────────────────────────────
length_cols <- c("length", "length_class", "ind_weight", "wgt", "wgt_cpue",
                 "num", "num_cpue")
present <- length_cols[length_cols %in% names(q1_83)]
absent  <- length_cols[!length_cols %in% names(q1_83)]

cat("Columns present:", paste(present, collapse = ", "), "\n")
cat("Columns absent: ", paste(absent,  collapse = ", "), "\n\n")
```

```{r weight-coverage}
cat("=== Weight coverage ===\n\n")

# wgt: total weight per species per haul
if ("wgt" %in% names(q1_83)) {
  cat("wgt (total catch weight per species x haul):\n")
  cat("  Non-NA rows:    ", sum(!is.na(q1_83$wgt)), "\n")
  cat("  NA rows:        ", sum( is.na(q1_83$wgt)), "\n")
  cat("  % non-NA:       ", round(100 * mean(!is.na(q1_83$wgt)), 1), "\n")
  cat("  Zero values:    ", sum(q1_83$wgt == 0, na.rm = TRUE), "\n")
  cat("  Positive values:", sum(q1_83$wgt  > 0, na.rm = TRUE), "\n\n")
}

# wgt_cpue: weight per unit effort
if ("wgt_cpue" %in% names(q1_83)) {
  cat("wgt_cpue (weight CPUE per species x haul):\n")
  cat("  Non-NA rows:    ", sum(!is.na(q1_83$wgt_cpue)), "\n")
  cat("  NA rows:        ", sum( is.na(q1_83$wgt_cpue)), "\n")
  cat("  % non-NA:       ", round(100 * mean(!is.na(q1_83$wgt_cpue)), 1), "\n")
  cat("  Zero values:    ", sum(q1_83$wgt_cpue == 0, na.rm = TRUE), "\n")
  cat("  Positive values:", sum(q1_83$wgt_cpue  > 0, na.rm = TRUE), "\n\n")
}

# Coverage by year — are there years with poor wgt coverage?
if ("wgt_cpue" %in% names(q1_83)) {
  cat("wgt_cpue coverage by year (% non-NA rows per year):\n")
  q1_83 %>%
    group_by(year) %>%
    summarise(
      n_rows      = n(),
      pct_wgt_ok  = round(100 * mean(!is.na(wgt_cpue)), 1),
      .groups     = "drop"
    ) %>%
    filter(pct_wgt_ok < 95) %>%
    { if (nrow(.) == 0) cat("  All years >= 95% coverage\n") else print(knitr::kable(., caption = "Years with < 95% wgt_cpue coverage")) }
}
```

```{r length-coverage}
cat("=== Length data coverage ===\n\n")

# Check for any length column
length_present <- intersect(c("length", "length_class"), names(q1_83))

if (length(length_present) == 0) {
  cat("No length or length_class column found in this dataset.\n")
  cat("Individual-level E calculation from length data is not feasible.\n")
  cat("E must be derived from species-mean body mass via FishBase.\n\n")
} else {
  for (col in length_present) {
    cat(col, ":\n")
    cat("  Non-NA rows:", sum(!is.na(q1_83[[col]])), "\n")
    cat("  NA rows:    ", sum( is.na(q1_83[[col]])), "\n")
    cat("  % non-NA:   ", round(100 * mean(!is.na(q1_83[[col]])), 1), "\n")
    if (is.numeric(q1_83[[col]])) {
      cat("  Range:      ", min(q1_83[[col]], na.rm=TRUE), "–",
                            max(q1_83[[col]], na.rm=TRUE), "\n")
    } else {
      cat("  Unique values (first 10):",
          paste(head(sort(unique(q1_83[[col]])), 10), collapse=", "), "\n")
    }
    cat("\n")
  }
}

# ind_weight: per-individual weight if present
if ("ind_weight" %in% names(q1_83)) {
  cat("ind_weight (per-individual weight):\n")
  cat("  Non-NA rows:", sum(!is.na(q1_83$ind_weight)), "\n")
  cat("  % non-NA:   ", round(100 * mean(!is.na(q1_83$ind_weight)), 1), "\n\n")
}
```

```{r abundance-coverage}
cat("=== Abundance coverage (num / num_cpue) ===\n\n")

if ("num_cpue" %in% names(q1_83)) {
  cat("num_cpue (abundance CPUE per species x haul):\n")
  cat("  Non-NA rows:    ", sum(!is.na(q1_83$num_cpue)), "\n")
  cat("  % non-NA:       ", round(100 * mean(!is.na(q1_83$num_cpue)), 1), "\n")
  cat("  Positive values:", sum(q1_83$num_cpue > 0, na.rm = TRUE), "\n\n")
}

# Rows where both wgt_cpue and num_cpue are positive — usable for EEoS
if (all(c("wgt_cpue", "num_cpue") %in% names(q1_83))) {
  both_ok <- sum(q1_83$wgt_cpue > 0 & q1_83$num_cpue > 0, na.rm = TRUE)
  cat("Rows with both wgt_cpue > 0 AND num_cpue > 0 (fully usable for B and N):\n")
  cat(" ", both_ok, "rows =",
      round(100 * both_ok / nrow(q1_83), 1), "% of Q1 >= 1983 data\n\n")
}

# Species with zero wgt_cpue in all hauls — potential B reliability issue
if ("wgt_cpue" %in% names(q1_83)) {
  cat("Species where wgt_cpue is ALWAYS zero or NA (no usable weight data):\n")
  species_wgt <- q1_83 %>%
    filter(!is.na(accepted_name)) %>%
    group_by(accepted_name) %>%
    summarise(
      n_records    = n(),
      n_wgt_ok     = sum(wgt_cpue > 0, na.rm = TRUE),
      pct_wgt_ok   = round(100 * n_wgt_ok / n_records, 1),
      .groups      = "drop"
    ) %>%
    arrange(pct_wgt_ok)
  
  no_wgt <- species_wgt %>% filter(n_wgt_ok == 0)
  cat("  Species with 0% weight coverage:", nrow(no_wgt), "\n")
  if (nrow(no_wgt) > 0) print(knitr::kable(head(no_wgt, 20)))
  
  cat("\nSpecies with < 50% weight coverage (may affect B reliability):\n")
  poor_wgt <- species_wgt %>% filter(pct_wgt_ok > 0, pct_wgt_ok < 50)
  cat("  Count:", nrow(poor_wgt), "\n")
  if (nrow(poor_wgt) > 0) print(knitr::kable(head(poor_wgt, 20)))
}
```

---

## FishBase coverage — E calculation species audit

Evaluates which of the ~245 NS-IBTS species have usable length–weight parameters
in FishBase, which are missing, and what proportion of total abundance they represent.
Requires the `rfishbase` package (`install.packages("rfishbase")`).

```{r fishbase-species-list}
library(dplyr)
library(knitr)

# ── Step 1: species list from Q1 >= 1983 panel ────────────────────────────────
# Use the consistent working species set: num > 0 AND wgt > 0
species_list <- data %>%
  filter(quarter == 1, year >= 1983,
         !is.na(accepted_name),
         num > 0, wgt > 0) %>%
  group_by(accepted_name, SpecCode) %>%
  summarise(
    total_abundance = sum(num,       na.rm = TRUE),
    total_records   = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(total_abundance))

cat("Total species in working set:", nrow(species_list), "\n")
cat("Total individual records:    ", sum(species_list$total_records), "\n\n")

# Proportion of total abundance per species
species_list <- species_list %>%
  mutate(pct_abundance = round(100 * total_abundance / sum(total_abundance), 3))

kable(head(species_list, 20),
      caption = "Top 20 species by total abundance in Q1 >= 1983 panel")
```

```{r fishbase-lw-lookup}
library(rfishbase)

cat("=== FishBase length-weight parameter lookup ===\n\n")

# ── Step 2: query length-weight parameters ────────────────────────────────────
# Use SpecCode (FishBase species code) for direct lookup where available
# Fall back to accepted_name for species with missing SpecCode

species_names <- species_list$accepted_name

# Query length_weight table from FishBase
lw_params <- tryCatch(
  length_weight(species_names, fields = c("Species", "a", "b", "Type", "Number")),
  error = function(e) { cat("rfishbase error:", conditionMessage(e), "\n"); NULL }
)

if (!is.null(lw_params) && nrow(lw_params) > 0) {

  # Keep only standard length-weight relationships (Type = "TL" or "Total")
  # and summarise to one row per species (best-available params)
  lw_summary <- lw_params %>%
    filter(!is.na(a), !is.na(b), b > 1, b < 5) %>%   # plausible b range
    group_by(Species) %>%
    summarise(
      n_lw_records = n(),
      a_median     = round(median(a, na.rm = TRUE), 6),
      b_median     = round(median(b, na.rm = TRUE), 3),
      .groups      = "drop"
    )

  cat("Species with usable L-W params:", nrow(lw_summary), "of", nrow(species_list), "\n\n")

  # ── Step 3: join back to species list ─────────────────────────────────────────
  species_audit <- species_list %>%
    left_join(lw_summary, by = c("accepted_name" = "Species")) %>%
    mutate(lw_available = !is.na(a_median))

  # ── Step 4: coverage summary ──────────────────────────────────────────────────
  n_with    <- sum(species_audit$lw_available)
  n_without <- sum(!species_audit$lw_available)
  pct_abund_covered <- round(
    100 * sum(species_audit$total_abundance[species_audit$lw_available]) /
          sum(species_audit$total_abundance), 1)

  cat("=== Coverage summary ===\n")
  cat("Species WITH L-W params:    ", n_with,    "(", round(100*n_with/nrow(species_audit),1), "%)\n")
  cat("Species WITHOUT L-W params: ", n_without, "(", round(100*n_without/nrow(species_audit),1), "%)\n")
  cat("% total abundance covered:  ", pct_abund_covered, "%\n\n")

  # ── Step 5: missing species ranked by abundance ───────────────────────────────
  missing <- species_audit %>%
    filter(!lw_available) %>%
    arrange(desc(total_abundance)) %>%
    select(accepted_name, total_abundance, pct_abundance, total_records)

  cat("Species missing L-W params, ranked by abundance:\n")
  kable(missing, caption = "Species without FishBase L-W parameters")

  # ── Step 6: cumulative abundance impact ───────────────────────────────────────
  cat("\n=== Cumulative abundance of missing species ===\n")
  cat("Top 5 missing species account for",
      round(sum(head(missing$pct_abundance, 5)), 2), "% of total abundance\n")
  cat("All missing species account for  ",
      round(100 - pct_abund_covered, 2), "% of total abundance\n\n")

  # ── Step 7: taxonomic fallback candidates ─────────────────────────────────────
  cat("=== Taxonomic fallback assessment ===\n")
  cat("For species missing L-W params, check if congeners have params:\n\n")

  missing_genera <- missing %>%
    mutate(genus = word(accepted_name, 1)) %>%
    left_join(
      species_audit %>%
        filter(lw_available) %>%
        mutate(genus = word(accepted_name, 1)) %>%
        group_by(genus) %>%
        summarise(n_congeners_with_lw = n(), .groups = "drop"),
      by = "genus"
    ) %>%
    select(accepted_name, genus, total_abundance, pct_abundance,
           n_congeners_with_lw) %>%
    mutate(fallback = case_when(
      !is.na(n_congeners_with_lw) ~ paste0("Genus mean (", n_congeners_with_lw, " congeners)"),
      TRUE ~ "Family mean required"
    ))

  kable(missing_genera,
        caption = "Missing species with taxonomic fallback assessment")

} else {
  cat("No L-W parameters returned. Check rfishbase connection or species name matching.\n")
  cat("Try: validate_names(species_names) to check name harmonisation.\n")
}
```

```{r fishbase-name-validation}
# ── Optional: validate species name matching ──────────────────────────────────
# Run this if the lookup above returns fewer matches than expected
# FishBase uses specific name formats that may differ from WoRMS-harmonised names

cat("=== Name validation (first 10 species) ===\n")
test_names <- head(species_list$accepted_name, 10)

validated <- tryCatch(
  validate_names(test_names),
  error = function(e) { cat("Validation error:", conditionMessage(e), "\n"); NULL }
)

if (!is.null(validated)) {
  name_check <- data.frame(
    original  = test_names,
    validated = validated
  ) %>%
    mutate(match = original == validated)
  kable(name_check, caption = "Name validation — original vs FishBase")
  cat("\nMismatches:", sum(!name_check$match), "of", nrow(name_check), "\n")
  cat("If mismatches are common, use SpecCode-based lookup instead.\n")
}
```

---

## Standardized file (optional)

FishGlob also ships **`NS-IBTS_std_clean.RData`** (standardized metrics). To inspect it in a new chunk:

```r
# p <- sub("NS-IBTS_clean", "NS-IBTS_std_clean", path_rdata)
# load(p)
# # same object names: data, readme
```

---

*Source: [FishGlob_data / outputs / Cleaned_data](https://github.com/fishglob/FishGlob_data/tree/main/outputs/Cleaned_data)*
