---
title: "H1 — Are there two trends in the main scatter?"
subtitle: "Quick diagnostic: residual distribution + decade colouring"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 5)

library(dplyr)
library(ggplot2)
library(kableExtra)

# Inline log_r2 — no sourcing needed
log_r2 <- function(log_obs, log_pred) {
  ss_res <- sum((log_obs - log_pred)^2, na.rm = TRUE)
  ss_tot <- sum((log_obs - mean(log_obs, na.rm = TRUE))^2, na.rm = TRUE)
  1 - ss_res / ss_tot
}

# Load pre-computed haul-level data (one RDS, ~1.5 MB, no pipeline needed)
data_path <- file.path(dirname(knitr::current_input(dir = TRUE)), "..",
                       "outputs", "haul_h1_benchmarks.rds")
haul <- readRDS(normalizePath(data_path, mustWork = TRUE))

n_hauls    <- nrow(haul)
r2_overall <- round(log_r2(haul$ln_B_obs, haul$ln_B_pred), 3)
med_resid  <- round(median(haul$residual, na.rm = TRUE), 3)
mean_resid <- round(mean(haul$residual,   na.rm = TRUE), 3)
```

---

## The question

The main scatter plot (log B_obs vs log B_pred) looked as though it might contain **two separate trends** rather than one diffuse cloud. This document checks:

1. **Is the residual distribution unimodal or bimodal?** — a true second group would leave a second peak in the residuals.
2. **Do the apparent trends separate by decade?** — whether any temporal structure is present in the haul-level residuals.

Overall log_r2 across all `r format(n_hauls, big.mark=",")` hauls: **`r r2_overall`**. Negative = EEoS overpredicts on average.

---

## 1. Residual distribution — one failure mode or two?

A negative residual means EEoS overpredicts (log B_pred > log B_obs). The median residual is **`r med_resid`** log-units (≈ `r round(exp(abs(med_resid)), 1)`× overprediction on the raw scale). If the distribution has a single peak, the "two trends" appearance is a density artifact from overplotting. A shoulder or second peak means a real subpopulation behaves differently.

```{r fig-resid-density, fig.cap="EEoS log-scale residuals. Blue vertical line = median; red dotted = mean. Negative = EEoS overpredicts. The rug along the bottom shows every individual haul.", fig.width=10, fig.height=5}
ggplot(haul, aes(x = residual)) +
  geom_histogram(aes(y = after_stat(density)),
    bins = 80, fill = "steelblue", colour = "white", alpha = 0.65) +
  geom_density(colour = "firebrick", linewidth = 1.0) +
  geom_rug(alpha = 0.04, colour = "grey40") +
  geom_vline(xintercept = 0,          linetype = 2, colour = "grey40",   linewidth = 0.8) +
  geom_vline(xintercept = med_resid,  linetype = 1, colour = "steelblue",linewidth = 1.0) +
  geom_vline(xintercept = mean_resid, linetype = 3, colour = "firebrick", linewidth = 1.0) +
  annotate("text", x = 0.12, y = Inf, vjust = 2, hjust = 0, size = 3.2, colour = "grey40",
           label = "0 = perfect") +
  annotate("text", x = med_resid - 0.1, y = Inf, vjust = 4, hjust = 1, size = 3.2,
           colour = "steelblue",
           label = paste0("median = ", med_resid)) +
  labs(
    x = "log-scale residual  [log(B_obs) \u2212 log(B_pred)]",
    y = "Density",
    title = "Are there one or two failure modes?",
    subtitle = "Unimodal = one failure mode (systematic overprediction)  |  Bimodal = two distinct haul populations"
  ) +
  theme_minimal(base_size = 12)
```

---

## 2. Main scatter coloured by decade

If any temporal structure exists in the haul-level residuals, it should be visible as a separation of decade colours within the cloud. Decades well-mixed throughout the cloud indicate no temporal structure at this scale — any such structure, if it exists, would need to emerge at aggregated spatial or temporal scales (H2, H3).

```{r fig-scatter-decade, fig.cap="**Supplementary Figure S2.** log(B_obs) vs log(B_pred) coloured by decade. Decades are well-mixed throughout the cloud with no systematic temporal separation at haul level. Whether residuals carry temporal structure at aggregated spatial scales is addressed in H2 and H3.", fig.width=11, fig.height=5.5}
haul_col <- haul |>
  mutate(decade = case_when(
    year <= 1994 ~ "1985\u20131994",
    year <= 2004 ~ "1995\u20132004",
    TRUE         ~ "2005\u20132015"
  ),
  decade = factor(decade, levels = c("1985\u20131994","1995\u20132004","2005\u20132015")))

ggplot(haul_col, aes(x = ln_B_obs, y = ln_B_pred, colour = decade)) +
  geom_point(alpha = 0.18, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey30", linewidth = 0.8) +
  scale_colour_manual(
    values = c("1985\u20131994" = "#2166ac",
               "1995\u20132004" = "#f4a582",
               "2005\u20132015" = "#d6604d"),
    name = "Decade"
  ) +
  labs(
    x = "log(B_obs) [g]",
    y = "log(B_pred) [g]",
    title = "Do the two apparent trends separate by decade?",
    subtitle = "1:1 line = perfect prediction — colours show temporal grouping"
  ) +
  guides(colour = guide_legend(override.aes = list(alpha = 0.9, size = 2.5))) +
  theme_minimal(base_size = 12)
```

---

## 3. Per-decade summary table

```{r decade-table}
decade_summary <- haul_col |>
  group_by(decade) |>
  summarise(
    n              = n(),
    log_r2         = round(log_r2(ln_B_obs, ln_B_pred), 3),
    median_residual = round(median(residual, na.rm = TRUE), 3),
    raw_overpredict = round(median(B_pred / B_obs, na.rm = TRUE), 1),
    .groups = "drop"
  )

decade_summary |>
  kable(
    col.names = c("Decade", "n hauls", "log_r2",
                  "Median log residual", "Median B_pred/B_obs"),
    caption = "EEoS performance by decade — does the failure mode change over time?"
  ) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover")) |>
  column_spec(3, bold = TRUE)
```

---

## 4. What to take from this

**Residual density (Figure S1):** Unimodal — a single, consistent failure mode. EEoS overpredicts with a roughly log-normal spread. There is no genuine second population of hauls with near-accurate prediction.

**Decade colouring (Figure S2):** Decades are well-mixed throughout the cloud. There is no systematic temporal separation at haul level. Whether aggregated spatial-temporal structure exists in EEoS residuals is the question H2 and H3 address.

**The apparent "two trends":** The cluster of points that appears visually closer to the 1:1 line is the high-biomass (Q4) tail of the distribution — large hauls have a smaller absolute prediction gap, which reads as a tighter group in a log-log scatter. This is a density artifact, not a distinct subpopulation. The unimodal residual distribution confirms this.
