---
title: "Hypothesis 1 — Supervisor briefing"
subtitle: "NS-IBTS Q1 (1985–2015) · EEoS haul-level biomass"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 5)
workspace_root <- if (file.exists("FishGlob_data")) {
  normalizePath(getwd(), winslash = "/", mustWork = TRUE)
} else if (file.exists("../FishGlob_data")) {
  normalizePath("..", winslash = "/", mustWork = TRUE)
} else {
  inp <- knitr::current_input(dir = TRUE)
  normalizePath(file.path(inp, ".."), winslash = "/", mustWork = TRUE)
}
r_dir <- file.path(workspace_root, "pipeline", "R")
source(file.path(r_dir, "h1_common.R"))
source(file.path(r_dir, "h1_lne_helpers.R"))
source(file.path(r_dir, "h1_null_helpers.R"))
project_root <- workspace_root
library(dplyr)
library(ggplot2)
library(kableExtra)
library(patchwork)

path_preds <- file.path(project_root, "outputs", "haul_eeos_predictions.rds")
path_benchmarks <- file.path(project_root, "outputs", "haul_h1_benchmarks.rds")
path_compare <- file.path(project_root, "outputs", "h1_model_comparison.rds")
path_null <- file.path(project_root, "outputs", "null_summary.rds")
path_dropout <- file.path(project_root, "outputs", "h1_dropout_summary.csv")
path_dropout_year <- file.path(project_root, "outputs", "h1_dropout_by_year.csv")
path_build_diag <- file.path(project_root, "outputs", "datras_build_diagnostics.csv")

stopifnot(file.exists(path_preds))
haul <- if (file.exists(path_benchmarks)) readRDS(path_benchmarks) else readRDS(path_preds)
compare <- if (file.exists(path_compare)) readRDS(path_compare) else NULL
null_sum <- if (file.exists(path_null)) readRDS(path_null) else NULL
decision <- assess_b_null_sampling(haul$B_obs)
dropout <- read.csv(path_dropout, stringsAsFactors = FALSE)
dropout_year <- read.csv(path_dropout_year, stringsAsFactors = FALSE)
build_diag <- read.csv(path_build_diag, stringsAsFactors = FALSE)

n_hauls <- nrow(haul)
log_B <- log(haul$B_obs)
log_B_eeos <- log(haul$B_pred)
r2_eeos <- log_r2(log_B, log_B_eeos)
cor2_eeos <- log_cor2(log_B, log_B_eeos)
med_ratio <- median(haul$B_pred / haul$B_obs, na.rm = TRUE)
med_abs_eeos <- median(haul$abs_residual, na.rm = TRUE)
r2_lne <- if (!is.null(compare)) {
  compare$comparison$log_r2[compare$comparison$model_id == "ln_e_ols"]
} else NA_real_
med_abs_lne <- if ("abs_residual_lnE" %in% names(haul)) median(haul$abs_residual_lnE, na.rm = TRUE) else NA_real_
ss_ratio <- if (!is.null(compare)) compare$ss_ratio_eeos_over_lne else NA_real_
harte_met <- if (!is.null(compare)) compare$harte_criterion_met_vs_lne else NA
n_datras <- as.integer(build_diag$value[build_diag$metric == "n_hauls"])
```

See [`pipeline/README.md`](../pipeline/README.md) for run order.

---

**H1:** EEoS prediction of haul-level community biomass (from S, N, E) outperforms both a B-only null model (random biomass assignment) and the best correlative alternative (ln(E) alone, Harte et al. 2022 benchmark).

Two tests:

| Test | Criterion | Result |
|------|-----------|--------|
| vs null (random B) | EEoS log_r2 > null distribution | **Passes** — p = `r if (!is.null(null_sum)) format.pval(null_sum$primary$p_value_one_sided, digits = 2, eps = 0.001) else "—"` |
| vs ln(E) correlative | EEoS SS_res < ½ × ln(E) SS_res (Harte) | **`r ifelse(isTRUE(harte_met), "Passes", "Fails")`** — SS ratio `r if (!is.null(ss_ratio)) sprintf("%.1f×", ss_ratio) else "—"` |

---

# Part A — Methods (notes)

## A1. Data and analysis sample

- **Survey:** NS-IBTS Quarter 1, 1985–2015
- **Observed biomass (`B_obs`):** FishGlob `sum(wgt) × 1000` → grams per haul ([`pipeline/build_eeos_predictions.R`](../pipeline/build_eeos_predictions.R))
- **State variables (S, N, E):** DATRAS HL length bins + FishBase LW lookup ([`pipeline/build_datras_state_variables.R`](../pipeline/build_datras_state_variables.R))
- **Join:** Inner join FishGlob ↔ DATRAS on 6-field haul key (Survey, Year, Quarter, Country, Ship, HaulNo)
- **Analysis panel:** **`r format(n_hauls, big.mark = ",")`** hauls — `r dropout$n_hauls[dropout$stage == "matched_datras_state"]` joined → `r dropout$n_hauls[dropout$stage == "failed_eeos_filters"]` excluded by EEoS filters → `r dropout$n_hauls[dropout$stage == "eeos_predictions"]` with predictions
- **Filter exclusions:** `B_obs > 0`, `S ≥ 2`, `E > N`, finite normalised E (44 zero-biomass FishGlob hauls, mostly 1991; 4 single-species hauls)
- **LW gaps:** Individuals in bins without usable FishBase a/b parameters are dropped from S, N, E (32 MISSING-fallback taxa in lookup; not imputed)

## A2. E calculation — key decisions

| Step | Decision |
|------|----------|
| Taxa | Fish only; cephalopods, crustaceans, bivalves excluded by AphiaID ([`pipeline/R/datras_constants.R`](../pipeline/R/datras_constants.R)) |
| Length units | LngtCode → cm midpoints (codes 0, 1, 5, `.`) ([`pipeline/R/datras_hl_helpers.R`](../pipeline/R/datras_hl_helpers.R)) |
| Mass | W(g) = a·L(cm)^b from frozen [`outputs/fishbase_lw_lookup_v2.csv`](outputs/fishbase_lw_lookup_v2.csv) (TL-type hierarchy, built offline) |
| LW gaps | Bins with missing/non-finite a/b or mass **dropped** before aggregation |
| N | `sum(HLNoAtLngt × SubFactor)` over LW-valid bins (same rows as E) |
| S | `n_distinct(AphiaID)` over same bins (= `n_species_with_lw`) |
| E_raw | `sum(N_at_bin × mass^0.75)` |
| m_min | Mass of **smallest length bin** in haul (not species mean length) |
| E (EEoS input) | `E = E_raw / m_min^0.75` so minimum individual metabolic rate = 1 |

## A3. EEoS prediction at haul level

- Per haul: **S, N, E** (normalised) → Python `biomass.biomass()` via reticulate ([`equation_of_state/biomass.py`](equation_of_state/biomass.py))
- Output: dimensionless `B_pred_norm`; **`B_pred = B_pred_norm × m_min`** (grams)
- Primary residual: **`log(B_obs) − log(B_pred)`** (negative ⇒ EEoS overpredicts catch biomass)
- Primary R²: **`log_r2()`** = 1 − SS_res/SS_tot (can be negative); **not** cor(log B)²

## A4. Correlative benchmark and null model

**ln(E) reference (Harte et al. 2022):**

- `lm(log(B_obs) ~ log(E))` on same hauls, same normalised E as EEoS ([`pipeline/run_h1_lne_reference.R`](../pipeline/run_h1_lne_reference.R))
- Harte criterion: EEoS unexplained SS should be **< ½** ln(E) SS

**B-only null (S/N/E not shuffled — breaks EEoS constraints):**

- Permuting S, N, or E violates hard constraints (`E > N`, integer counts, biological hierarchy); only `B_obs` is randomised ([`pipeline/run_h1_null_model.R`](../pipeline/run_h1_null_model.R))
- Shape of `log(B_obs)` assessed automatically ([`pipeline/R/h1_null_helpers.R`](../pipeline/R/h1_null_helpers.R)): **`r decision$rationale`** → uniform null preferred over B shuffle (peaked empirical distribution makes shuffle too lenient)
- **Primary null:** uniform draw from central 95% of log(B_obs) (`uniform_log_b_95`; 999 permutations, seed 42)
- **Robustness null:** permute observed B_obs across hauls (`b_shuffle`)
- **Test statistic:** one-sided p = P(null log_r2 ≥ observed log_r2)

---

# Part B — Results

## Key numbers

```{r key-numbers}
key_tbl <- tibble(
  Quantity = c(
    "Analysis hauls (n)",
    "Year range",
    "Median B_obs (g)",
    "Median B_pred (g)",
    "Median B_pred / B_obs",
    "EEoS log_r2",
    "ln(E) log_r2",
    "SS_res ratio (EEoS / ln(E))",
    "Harte criterion met?",
    "cor(log B)² (diagnostic only)",
    "Primary null p (one-sided)",
    "Robustness null p (one-sided)",
    "Median |residual| EEoS",
    "Median |residual| ln(E)"
  ),
  Value = c(
    format(n_hauls, big.mark = ","),
    "1985–2015",
    format(round(median(haul$B_obs)), big.mark = ","),
    format(round(median(haul$B_pred)), big.mark = ","),
    sprintf("%.2f", med_ratio),
    sprintf("%.3f", r2_eeos),
    sprintf("%.3f", r2_lne),
    sprintf("%.2f×", ss_ratio),
    ifelse(isTRUE(harte_met), "Yes", "No"),
    sprintf("%.3f", cor2_eeos),
    if (!is.null(null_sum)) format.pval(null_sum$primary$p_value_one_sided, digits = 3, eps = 0.001) else "—",
    if (!is.null(null_sum)) format.pval(null_sum$robustness$p_value_one_sided, digits = 3, eps = 0.001) else "—",
    sprintf("%.3f", med_abs_eeos),
    sprintf("%.3f", med_abs_lne)
  )
)
kable(key_tbl, col.names = c("Quantity", "Value"), caption = "H1 summary statistics") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"), position = "left")
```

## Figure 1 — Headline (H1 performance + Harte comparison)

```{r fig-headline, fig.cap="Figure 1. Panel A: observed vs EEoS predicted biomass (log scale). Panel B: log-scale predictive comparison (EEoS vs ln(E) reference).", fig.height=5, fig.width=11}
p_a <- ggplot(haul, aes(x = ln_B_obs, y = ln_B_pred)) +
  geom_point(alpha = 0.1, size = 0.6, colour = "grey30") +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick", linewidth = 0.9) +
  annotate(
    "label", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5, size = 3.2,
    label = sprintf(
      "n = %s\nlog_r2 = %.2f\ncor² = %.2f\nmedian B_pred/B_obs = %.1f×",
      format(n_hauls, big.mark = ","), r2_eeos, cor2_eeos, med_ratio
    ),
    fill = "white", alpha = 0.85, label.size = 0
  ) +
  labs(
    x = expression(log(B[obs] ~ "(g)")),
    y = expression(log(B[pred] ~ "(g)")),
    title = "A  EEoS vs observed biomass"
  ) +
  theme_minimal(base_size = 11)

cmp_df <- if (!is.null(compare)) {
  compare$comparison |>
    filter(model_id %in% c("eeos_biomass", "ln_e_ols")) |>
    mutate(model = factor(model, levels = model))
} else {
  tibble(model = character(), log_r2 = numeric(), ss_res_log = numeric())
}

p_b <- ggplot(cmp_df, aes(x = model, y = log_r2, fill = model)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = 3, colour = "grey40") +
  scale_fill_manual(values = c(
    "EEoS (S, N, E)" = "firebrick",
    "ln(E) OLS (fitted correlative)" = "darkgreen"
  )) +
  annotate(
    "text", x = 1.5, y = max(cmp_df$log_r2, 0) + 0.08,
    label = sprintf("SS ratio = %.1f×\nHarte vs ln(E): %s", ss_ratio, ifelse(isTRUE(harte_met), "met", "NOT met")),
    size = 3.2
  ) +
  labs(
    x = NULL,
    y = expression(log~scale~R^2),
    title = "B  Predictive comparison (Harte criterion)"
  ) +x=
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

p_a + p_b + plot_layout(widths = c(1.15, 1))
```

**Panel A — how to read it.** Points above the 1:1 line indicate B_pred > B_obs (EEoS overpredicts). The tight diagonal cloud with high cor² (`r sprintf("%.2f", cor2_eeos)`) reflects **rank association** — hauls with larger predicted biomass tend to have larger observed biomass. However the cloud sits systematically above the 1:1 line, and **log_r2 = `r sprintf("%.2f", r2_eeos)`** means EEoS predictions are worse than simply predicting the mean log(B_obs) for every haul. Association ≠ prediction.

**Panel B — how to read it.** ln(E) alone (green) achieves a positive log_r2 (`r sprintf("%.2f", r2_lne)`); EEoS (red) is below zero (`r sprintf("%.2f", r2_eeos)`). Adding S and N through the equation of state does not improve on metabolic rate alone — EEoS unexplained SS is `r sprintf("%.1f×", ss_ratio)` that of ln(E), so the **Harte criterion is `r ifelse(isTRUE(harte_met), "met", "not met")`**. H1 `r ifelse(isTRUE(harte_met), "passes", "fails")` the correlative comparison.

## Figure 2 — Null model (does EEoS beat random chance?)

**Why randomise B_obs, not S/N/E?** EEoS state variables must satisfy hard mathematical constraints: `E > N`, integer counts, and the biological hierarchy `N >> S`, `E >> N`. Permuting S, N, or E across hauls routinely violates these constraints and produces invalid inputs to the equation of state. The only quantity that can be freely randomised while keeping `B_pred` unchanged is **`B_obs`** (`pipeline/run_h1_null_model.R`).

**Why a uniform null, not a simple B shuffle?** The choice depends on the shape of `log(B_obs)`. An automated assessment (`pipeline/R/h1_null_helpers.R`) flags three criteria:

| Metric | Value | Peaked threshold |
|--------|-------|-----------------|
| IQR / range | **`r round(decision$metrics$iqr_range_ratio, 2)`** | < 0.25 |
| Histogram peak / uniform | **`r round(decision$metrics$hist_peak_ratio, 2)`** | > 2.5 |
| Excess kurtosis | **`r round(decision$metrics$excess_kurtosis, 2)`** | > 0.75 |

`r decision$rationale` A **B shuffle** preserves this peaked empirical distribution and would therefore be a lenient null — the shuffled biomass values would tend to cluster in the same narrow range as observed, making it easier for EEoS to "beat" the null. The **uniform null** (`uniform_log_b_95`) draws `B_obs` uniformly from the central 95% range of `log(B_obs)`, spreading randomised values more evenly and setting a harder baseline. The B shuffle is retained as a robustness check.

```{r fig-null, fig.cap="Figure 2. Primary null: distribution of log-scale R² under uniform B_obs randomisation (999 permutations). Red line = observed EEoS log_r2.", eval=!is.null(null_sum) && length(null_sum$figures) >= 1, fig.height=4.5, fig.width=8}
knitr::include_graphics(null_sum$figures[1])
```

```{r null-table, eval=!is.null(null_sum)}
null_tbl <- bind_rows(
  Primary = null_sum$primary,
  Robustness = null_sum$robustness,
  .id = "Role"
) |>
  transmute(
    Role,
    Scheme = if_else(Role == "Primary", "Uniform on 95% log(B)", "B shuffle"),
    `Observed log_r2` = sprintf("%.3f", null_sum$T_obs),
    `Null median log_r2` = sprintf("%.3f", null_median),
    `Null 95% interval` = sprintf("[%.3f, %.3f]", null_q025, null_q975),
    `p (one-sided)` = vapply(p_value_one_sided, function(p) format.pval(p, digits = 3, eps = 0.001), character(1))
  )
kable(null_tbl, caption = "Null-model comparison (B_pred fixed from observed S, N, E)") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

**Null result interpretation.** The observed log_r2 (`r sprintf("%.2f", r2_eeos)`) is substantially better than null medians under both schemes (`r if (!is.null(null_sum)) sprintf("%.2f", null_sum$primary$null_median) else "—"` for uniform, `r if (!is.null(null_sum)) sprintf("%.2f", null_sum$robustness$null_median) else "—"` for B shuffle), with p = `r if (!is.null(null_sum)) format.pval(null_sum$primary$p_value_one_sided, digits=2, eps=0.001) else "—"` one-sided (primary null). EEoS predictions carry **non-random co-structure** with observed biomass — there is genuine information in (S, N, E) about haul-level B. However this does not mean the predictions are *accurate*: log_r2 remains negative and EEoS consistently overpredicts (median B_pred/B_obs ≈ `r sprintf("%.1f×", med_ratio)`).

---

## Overall verdict

| Question | Test | Result |
|----------|------|--------|
| Does EEoS do better than random? | Null model (uniform B, p one-sided) | **`r ifelse(!is.null(null_sum) && null_sum$primary$p_value_one_sided < 0.05, "Yes", "No")`** — p = `r if (!is.null(null_sum)) format.pval(null_sum$primary$p_value_one_sided, digits=2, eps=0.001) else "—"` |
| Does EEoS beat the correlative benchmark? | Harte criterion (SS ratio vs ln(E)) | **`r ifelse(isTRUE(harte_met), "Yes", "No")`** — SS ratio `r if (!is.null(ss_ratio)) sprintf("%.1f×", ss_ratio) else "—"` |
| Is EEoS accurate in absolute terms? | log_r2 > 0 | **`r ifelse(r2_eeos > 0, "Yes", "No")`** — log_r2 = `r sprintf("%.2f", r2_eeos)` |

**H1 is half-supported.** EEoS captures non-random biomass co-structure (passes the null test) but does not outperform a simple metabolic-rate regression. The signal exists; the full equation of state does not translate it into better predictions than ln(E) alone.

---

## Supplementary — sample and coverage

### Sample funnel

```{r funnel}
funnel <- tibble(
  Stage = c(
    "DATRAS hauls with valid state (Q1 1985–2015)",
    "FishGlob ↔ DATRAS joined",
    "Excluded by EEoS filters",
    "EEoS predictions (analysis panel)"
  ),
  `N hauls` = c(
    n_datras,
    dropout$n_hauls[dropout$stage == "matched_datras_state"],
    dropout$n_hauls[dropout$stage == "failed_eeos_filters"],
    dropout$n_hauls[dropout$stage == "eeos_predictions"]
  )
)
kable(funnel, caption = "Haul counts at each pipeline stage") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

### Hauls by year

```{r fig-year, fig.cap="Analysis hauls by year (joined vs predicted). Dropout concentrated in 1991 (zero FishGlob biomass).", fig.height=4.5}
dropout_year |>
  mutate(year = factor(year)) |>
  ggplot(aes(x = year)) +
  geom_col(aes(y = n_joined), fill = "grey80", width = 0.75) +
  geom_col(aes(y = n_predicted), fill = "steelblue", width = 0.75) +
  labs(
    x = "Year",
    y = "Haul count",
    title = "Temporal coverage: joined (grey) vs predicted (blue)"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7))
```

### State variable ranges

```{r state-summary}
state_tbl <- tibble(
  Variable = c("S", "log(N)", "log(E)", "log(B_obs)", "m_min (g)"),
  Min = c(
    min(haul$S), min(log(haul$N)), min(log(haul$E)),
    min(log_B), min(haul$m_min)
  ),
  Median = c(
    median(haul$S), median(log(haul$N)), median(log(haul$E)),
    median(log_B), median(haul$m_min)
  ),
  Max = c(
    max(haul$S), max(log(haul$N)), max(log(haul$E)),
    max(log_B), max(haul$m_min)
  )
) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

kable(state_tbl, caption = "State variable ranges on analysis panel") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

```{r fig-state, fig.cap="Distributions of state variables and observed biomass.", fig.height=7, fig.width=10}
long_haul <- haul |>
  transmute(
    S = S,
    `log(N)` = log(N),
    `log(E)` = log(E),
    `log(B_obs)` = log_B
  ) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value")

ggplot(long_haul, aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white", alpha = 0.85) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(x = "Value", y = "Count", title = "State variable distributions") +
  theme_minimal(base_size = 10)
```

### log(B_obs) shape and null bounds

```{r fig-logb-null, fig.cap="Peaked log(B_obs) distribution with 2.5% and 97.5% quantiles (primary null bounds).", fig.height=4}
ggplot(data.frame(logB = log_B), aes(x = logB)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = decision$log_bounds_95, linetype = 2, colour = "red") +
  labs(
    x = expression(log(B[obs]) ~ "(g)"),
    y = "Haul count",
    title = "Observed haul biomass (log scale)",
    subtitle = "Red lines = uniform-null bounds (central 95%)"
  ) +
  theme_minimal()
```

### Residual distributions

```{r fig-residuals, fig.cap="Log-scale residual distributions: EEoS vs ln(E) reference.", fig.height=4.5, eval="residual_lnE" %in% names(haul)}
resid_long <- haul |>
  transmute(
    EEoS = residual,
    `ln(E)` = residual_lnE
  ) |>
  tidyr::pivot_longer(everything(), names_to = "model", values_to = "residual")

ggplot(resid_long, aes(x = residual, fill = model)) +
  geom_histogram(bins = 50, alpha = 0.55, position = "identity") +
  scale_fill_manual(values = c("EEoS" = "firebrick", "ln(E)" = "darkgreen")) +
  labs(x = "Log-scale residual", y = "Count", fill = NULL,
       title = "Residual distributions (negative = model overpredicts)") +
  theme_minimal()
```

### B_pred/B_obs vs community size

```{r fig-ratio-n, fig.cap="Median overprediction (B_pred/B_obs) vs log(N) by decile.", fig.height=4.5}
haul |>
  mutate(log_N = log(N), ratio = B_pred / B_obs) |>
  ggplot(aes(x = log_N, y = ratio)) +
  geom_point(alpha = 0.08, size = 0.7, colour = "grey40") +
  geom_smooth(method = "loess", se = TRUE, colour = "firebrick", linewidth = 0.9) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  scale_y_log10() +
  labs(
    x = expression(log(N)),
    y = "B_pred / B_obs",
    title = "Overprediction vs haul abundance"
  ) +
  theme_minimal()
```

### Metric clarification (log_r2 vs cor²)

```{r metric-clarify}
metric_tbl <- tibble(
  Metric = c("log_r2 (primary H1)", "cor(log B)² (association only)"),
  Value = c(sprintf("%.3f", r2_eeos), sprintf("%.3f", cor2_eeos)),
  Interpretation = c(
    "Predictive: 1 − SS_res/SS_tot; negative = worse than mean",
    "Rank association only; can be high despite poor prediction"
  )
)
kable(metric_tbl, caption = "Why cor² ≠ predictive success") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

### Spatial summary (ICES StatRec)

```{r stat-rec}
stat_rec_summary <- haul |>
  group_by(stat_rec) |>
  summarise(
    n = n(),
    median_ratio = median(B_pred / B_obs, na.rm = TRUE),
    median_residual = median(residual, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n >= 5)

top_over <- stat_rec_summary |> arrange(desc(median_ratio)) |> slice_head(n = 10)
top_under <- stat_rec_summary |> arrange(median_ratio) |> slice_head(n = 10)

bind_rows(
  top_over |> mutate(rank = "Highest B_pred/B_obs"),
  top_under |> mutate(rank = "Lowest B_pred/B_obs")
) |>
  mutate(
    median_ratio = round(median_ratio, 2),
    median_residual = round(median_residual, 2)
  ) |>
  select(rank, stat_rec, n, median_ratio, median_residual) |>
  kable(
    col.names = c("Group", "StatRec", "n hauls", "Median B_pred/B_obs", "Median residual"),
    caption = "ICES rectangles with strongest scale offset (n ≥ 5 hauls)"
  ) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

```{r fig-statrec, fig.cap="Distribution of median B_pred/B_obs across ICES StatRec rectangles (n ≥ 5 hauls).", fig.height=4}
ggplot(stat_rec_summary, aes(x = median_ratio)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = med_ratio, linetype = 2, colour = "red") +
  labs(
    x = "Median B_pred / B_obs per StatRec",
    y = "Number of rectangles",
    title = "Spatial heterogeneity in scale offset",
    subtitle = "Red line = overall median"
  ) +
  theme_minimal()
```
---