---
title: "H1 Null Model — Distribution Review & Sampling Decision"
subtitle: "North Sea EEoS haul-level biomass (NS-IBTS Q1)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 5)
workspace_root <- if (file.exists("FishGlob_data")) {
  normalizePath(getwd(), winslash = "/", mustWork = TRUE)
} else if (file.exists("../FishGlob_data")) {
  normalizePath("..", winslash = "/", mustWork = TRUE)
} else {
  inp <- knitr::current_input(dir = TRUE)
  normalizePath(file.path(inp, ".."), winslash = "/", mustWork = TRUE)
}
r_dir <- file.path(workspace_root, "pipeline", "R")
source(file.path(r_dir, "h1_common.R"))
source(file.path(r_dir, "h1_null_helpers.R"))
project_root <- workspace_root
library(dplyr)
library(ggplot2)
library(kableExtra)
```

## Purpose

Before running the H1 null model, inspect the shape of haul-level parameters — especially **`log(B_obs)`** — to decide how to randomise observed biomass without breaking EEoS constraints on **S**, **N**, and **E**.

Supervisor guidance: independently shuffling S/N/E is invalid (breaks `E > N`, `N >> S`, etc.). The null randomises **`B_obs`** while **`B_pred`** (from observed S, N, E) stays fixed.

## Load predictions

```{r load}
path_preds <- file.path(project_root, "outputs", "haul_eeos_predictions.rds")
stopifnot(file.exists(path_preds))
haul <- readRDS(path_preds)
logB <- log(haul$B_obs)
logBp <- log(haul$B_pred)
```

**`r nrow(haul)`** hauls with successful EEoS predictions.

## Distribution shape — decision inputs

```{r assess}
decision <- assess_b_null_sampling(haul$B_obs)
knitr::kable(decision$metrics, digits = 3, caption = "log(B_obs) shape metrics")
```

**Decision:** **`r decision$primary_method`** (primary); **`r decision$robustness_method`** (robustness check).

> **`r decision$rationale`**

### Criteria (automated)

| Criterion | Threshold | This data |
|-----------|-----------|-----------|
| IQR / range on log scale | `< 0.25` → peaked | **`r round(decision$metrics$iqr_range_ratio, 3)`** |
| Histogram peak / uniform | `> 2.5` → peaked | **`r round(decision$metrics$hist_peak_ratio, 2)`** |
| Excess kurtosis | `> 0.75` → peaked | **`r round(decision$metrics$excess_kurtosis, 2)`** |

If **any** criterion flags peaked → primary null = **uniform on central 95% of log(B_obs)**; otherwise primary = **B shuffle** (permute observed biomass across hauls).

## Histograms

```{r fig-log-b, fig.cap="log(B_obs) — primary null target"}
ggplot(data.frame(logB = logB), aes(x = logB)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = decision$log_bounds_95, linetype = 2, colour = "red") +
  labs(
    x = expression(log(B[obs]) ~ "(g)"),
    y = "Haul count",
    title = "Observed haul biomass (log scale)",
    subtitle = "Red dashed lines = 2.5th and 97.5th percentiles (uniform-null bounds if peaked)"
  )
```

```{r fig-obs-vs-pred, fig.cap="Observed vs EEoS predicted biomass (log-log)"}
ggplot(haul, aes(x = log(B_obs), y = log(B_pred))) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey40") +
  geom_smooth(method = "lm", se = FALSE, colour = "red") +
  labs(
    x = expression(log(B[obs])),
    y = expression(log(B[pred])),
    title = "EEoS underprediction at haul level"
  )
```

```{r fig-constraints, fig.cap="Why S/N/E cannot be shuffled independently"}
par(mfrow = c(1, 2))
hist(log(haul$N / haul$S), breaks = 40, main = "log(N/S)", xlab = "", col = "grey80")
hist(log(haul$E / haul$N), breaks = 40, main = "log(E/N)", xlab = "", col = "grey80")
par(mfrow = c(1, 1))
```

## Null model results

Run `Rscript run_h1_null_model.R` to generate permutations. Results below load saved output if present.

```{r null-results}
path_summary <- file.path(project_root, "outputs", "null_summary.rds")

if (!file.exists(path_summary)) {
  cat("Run `Rscript run_h1_null_model.R` to populate null_summary.rds\n")
} else {
  null_sum <- readRDS(path_summary)

  observed_tbl <- tibble(
    Statistic = c(
      "R² (log B_obs vs log B_pred)",
      "Median |residual| (log scale)"
    ),
    Observed = c(
      sprintf("%.3f", null_sum$T_obs),
      sprintf("%.3f", null_sum$median_abs_residual_obs)
    )
  )

  null_label <- function(method) {
    if (grepl("uniform", method, fixed = TRUE)) {
      "Primary — uniform on central 95% of log(B_obs)"
    } else if (grepl("shuffle", method, fixed = TRUE)) {
      "Robustness — B shuffle across hauls"
    } else {
      method
    }
  }

  null_tbl <- bind_rows(
    Primary = null_sum$primary,
    Robustness = null_sum$robustness,
    .id = "Role"
  ) |>
    transmute(
      Role,
      `Null scheme` = vapply(method, null_label, character(1)),
      `Null R² median` = sprintf("%.3f", null_median),
      `Null R² 95% interval` = sprintf(
        "[%.3f, %.3f]",
        null_q025,
        null_q975
      ),
      `p (one-sided)` = vapply(
        p_value_one_sided,
        function(p) format.pval(p, digits = 3, eps = 0.001),
        character(1)
      ),
      `p (two-sided)` = vapply(
        p_value_two_sided,
        function(p) format.pval(p, digits = 3, eps = 0.001),
        character(1)
      ),
      `Median |residual| (null)` = sprintf("%.3f", median_abs_residual_null)
    )

  cat("**Observed EEoS performance** (n = 999 permutations for null comparisons)\n\n")

  kable(
    observed_tbl,
    col.names = c("Statistic", "Observed value"),
    align = c("l", "r"),
    caption = "Observed haul-level fit"
  ) |>
    kable_styling(full_width = FALSE, position = "left") |>
    print()

  cat("\n\n**Null comparisons** — B_obs randomised; B_pred fixed from observed S, N, E\n\n")
  cat(
    "One-sided p = P(null R² ≥ observed R²). ",
    "Higher R² (less negative) favours the observed pairing.\n\n"
  )

  kable(
    null_tbl,
    col.names = c(
      "Role",
      "Null scheme",
      "Null R² median",
      "Null R² 95% interval",
      "p (one-sided)",
      "p (two-sided)",
      "Median |residual| (null)"
    ),
    align = c("l", "l", rep("r", 5)),
    caption = "Null-model R² on log scale (999 permutations per scheme)"
  ) |>
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |>
    column_spec(1, bold = TRUE) |>
    print()
}
```

## Checklist

```{r checklist}
checks <- data.frame(
  check = c(
    "log(B_obs) distribution assessed",
    "Primary null method chosen from shape criteria",
    "S/N/E not shuffled (B-only null)",
    "Null permutations computed (999)",
    "Observed R² compared to null distribution"
  ),
  status = c(
    "PASS",
    paste0("PASS (", decision$primary_method, ")"),
    "PASS",
    if (file.exists(path_summary)) "PASS" else "PENDING — run run_h1_null_model.R",
    if (file.exists(path_summary)) "PASS" else "PENDING"
  ),
  action_needed = c("—", "—", "—", "—", "—")
)
knitr::kable(checks)
```
