---
title: "Hypothesis 1 — EEoS haul-level biomass review"
subtitle: "NS-IBTS Q1 (1985–2015) · North Sea EEoS project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 5)
workspace_root <- if (file.exists("FishGlob_data")) {
  normalizePath(getwd(), winslash = "/", mustWork = TRUE)
} else if (file.exists("../FishGlob_data")) {
  normalizePath("..", winslash = "/", mustWork = TRUE)
} else {
  inp <- knitr::current_input(dir = TRUE)
  normalizePath(file.path(inp, ".."), winslash = "/", mustWork = TRUE)
}
r_dir <- file.path(workspace_root, "pipeline", "R")
source(file.path(r_dir, "h1_common.R"))
source(file.path(r_dir, "h1_lne_helpers.R"))
source(file.path(r_dir, "h1_null_helpers.R"))
project_root <- workspace_root
library(dplyr)
library(ggplot2)
library(kableExtra)
```

## Overview

**H1:** Does EEoS predict haul-level community biomass from state variables S, N, E?

**Pipeline** (see [`pipeline/README.md`](../pipeline/README.md)):

1. `Rscript pipeline/build_datras_state_variables.R`
2. `Rscript pipeline/build_eeos_predictions.R`
3. `Rscript pipeline/run_h1_lne_reference.R`
4. `Rscript pipeline/run_h1_null_model.R`

State variables **S, N, E** come from DATRAS HL (same filtered rows); **B_obs** from FishGlob.

Knit this report after steps 1–4 complete.

---

## Load results

```{r paths}
path_preds <- file.path(project_root, "outputs", "haul_eeos_predictions.rds")
path_benchmarks <- file.path(project_root, "outputs", "haul_h1_benchmarks.rds")
path_compare <- file.path(project_root, "outputs", "h1_model_comparison.rds")
path_null <- file.path(project_root, "outputs", "null_summary.rds")

stopifnot(file.exists(path_preds))
haul <- if (file.exists(path_benchmarks)) readRDS(path_benchmarks) else readRDS(path_preds)
compare <- if (file.exists(path_compare)) readRDS(path_compare) else NULL
null_sum <- if (file.exists(path_null)) readRDS(path_null) else NULL
decision <- assess_b_null_sampling(haul$B_obs)
```

**`r nrow(haul)`** hauls in analysis.

---

## 1. EEoS predictions

State variables **S, N, E** are assembled from DATRAS HL (+ FishBase LW); **B_obs** from FishGlob. Only **B_pred** is EEoS model output.

Primary residual: **`residual = log(B_obs) − log(B_pred)`** (both in grams).

```{r eeos-summary}
log_B <- log(haul$B_obs)
log_B_eeos <- log(haul$B_pred)
eeos_tbl <- tibble(
  Statistic = c(
    "R² (log B_obs vs log B_pred)",
    "Median |residual|",
    "Median B_pred / B_obs"
  ),
  Value = c(
    sprintf("%.3f", log_r2(log_B, log_B_eeos)),
    sprintf("%.3f", median(haul$abs_residual, na.rm = TRUE)),
    sprintf("%.3f", median(haul$B_pred / haul$B_obs, na.rm = TRUE))
  )
)
kable(eeos_tbl, col.names = c("Statistic", "Value"), caption = "EEoS haul-level fit") |>
  kable_styling(full_width = FALSE, position = "left")
```

```{r fig-eeos, fig.cap="Observed vs EEoS predicted biomass (log-log)"}
ggplot(haul, aes(x = log(B_obs), y = log(B_pred))) +
  geom_point(alpha = 0.12, size = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey45") +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  labs(
    x = expression(log(B[obs] ~ "(g)")),
    y = expression(log(B[pred] ~ "(g)")),
    title = "EEoS: systematic underprediction at haul level"
  )
```

---

## 2. ln(E) reference (Harte et al. 2022)

Single-predictor benchmark: **`lm(log(B_obs) ~ log(E))`** using the same normalised **E** passed to EEoS.

```{r lne-section}
if (is.null(compare)) {
  cat("Run `Rscript run_h1_lne_reference.R` to populate comparison outputs.\n")
} else {
  coef_tbl <- compare$coefficients
  coef_tbl$Estimate <- sprintf("%.4f", coef_tbl$Estimate)
  coef_tbl$`Std. Error` <- sprintf("%.4f", coef_tbl$`Std. Error`)
  coef_tbl$`t value` <- sprintf("%.2f", as.numeric(coef_tbl$`t value`))
  coef_tbl$`Pr(>|t|)` <- vapply(coef_tbl$`Pr(>|t|)`, function(p) format.pval(as.numeric(p), digits = 3, eps = 0.001), character(1))

  kable(
    coef_tbl,
    caption = "ln(E) reference regression coefficients"
  ) |>
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
}
```

```{r fig-lne, fig.cap="ln(E) reference: log(B_obs) vs log(E)"}
ggplot(haul, aes(x = log(E), y = log(B_obs))) +
  geom_point(alpha = 0.12, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE, colour = "darkgreen") +
  labs(
    x = expression(log(E) ~ "(normalised)"),
    y = expression(log(B[obs] ~ "(g)")),
    title = "ln(E) reference model"
  )
```

---

## 3. EEoS vs ln(E) comparison

Harte et al. (2022) criterion: EEoS unexplained variance should be **less than half** that of the second-best predictor (here, ln(E)).

```{r model-compare}
if (!is.null(compare)) {
  cmp <- compare$comparison |>
    mutate(
      `R² (log)` = sprintf("%.3f", log_r2),
      `SS_res (log)` = format(round(ss_res_log), big.mark = ","),
      `Median |residual|` = sprintf("%.3f", median_abs_log_resid),
      `SS ratio vs EEoS` = sprintf("%.2f×", ss_ratio_vs_eeos),
      `Harte vs ln(E)?` = ifelse(
        is.na(harte_criterion_vs_lne),
        "—",
        ifelse(harte_criterion_vs_lne, "Yes", "No")
      )
    ) |>
    select(
      Model = model,
      Tier = tier,
      `R² (log)`,
      `SS_res (log)`,
      `Median |residual|`,
      `SS ratio vs EEoS`,
      `Harte vs ln(E)?`
    )

  kable(cmp, caption = "Unified model comparison on log scale (same hauls)") |>
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |>
    column_spec(1, bold = TRUE)

  cat(
    "\n\n**Overall SS_res ratio (EEoS / ln(E) OLS):**",
    sprintf("%.2f×", compare$ss_ratio_eeos_over_lne),
    " — values > 1 mean ln(E) explains log biomass better than EEoS.\n"
  )
}
```

```{r fig-residuals-eeos, fig.cap="EEoS residual distribution (log scale)"}
ggplot(haul, aes(x = residual)) +
  geom_histogram(bins = 50, fill = "firebrick", alpha = 0.5) +
  labs(x = "EEoS residual", y = "Count") +
  theme_minimal()
```

```{r fig-residuals-lne, fig.cap="ln(E) reference residual distribution (log scale)", eval="residual_lnE" %in% names(haul)}
ggplot(haul, aes(x = residual_lnE)) +
  geom_histogram(bins = 50, fill = "darkgreen", alpha = 0.5) +
  labs(x = "ln(E) residual", y = "Count") +
  theme_minimal()
```

---

## 4. B-only null model

S/N/E are **not** shuffled (breaks EEoS constraints). **`B_obs`** is randomised while **`B_pred`** stays fixed.

Null scheme chosen from **`log(B_obs)`** shape (see [`run_h1_null_model.R`](run_h1_null_model.R)):

> **`r decision$rationale`**

```{r null-results}
if (is.null(null_sum)) {
  cat("Run `Rscript run_h1_null_model.R` to populate null outputs.\n")
} else {
  observed_tbl <- tibble(
    Statistic = c(
      "R² (log B_obs vs log B_pred)",
      "Median |residual| (log scale)"
    ),
    Observed = c(
      sprintf("%.3f", null_sum$T_obs),
      sprintf("%.3f", null_sum$median_abs_residual_obs)
    )
  )

  null_label <- function(method) {
    if (grepl("uniform", method, fixed = TRUE)) {
      "Primary — uniform on central 95% of log(B_obs)"
    } else if (grepl("shuffle", method, fixed = TRUE)) {
      "Robustness — B shuffle across hauls"
    } else {
      method
    }
  }

  null_tbl <- bind_rows(
    Primary = null_sum$primary,
    Robustness = null_sum$robustness,
    .id = "Role"
  ) |>
    transmute(
      Role,
      `Null scheme` = vapply(method, null_label, character(1)),
      `Null R² median` = sprintf("%.3f", null_median),
      `Null R² 95% interval` = sprintf("[%.3f, %.3f]", null_q025, null_q975),
      `p (one-sided)` = vapply(p_value_one_sided, function(p) format.pval(p, digits = 3, eps = 0.001), character(1)),
      `Median |residual| (null)` = sprintf("%.3f", median_abs_residual_null)
    )

  kable(observed_tbl, col.names = c("Statistic", "Observed value"),
        caption = "Observed EEoS performance") |>
    kable_styling(full_width = FALSE, position = "left")

  cat("\n\n")
  kable(null_tbl, caption = "Null-model R² (999 permutations per scheme)") |>
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |>
    column_spec(1, bold = TRUE)
}
```

```{r fig-null-primary, fig.cap="Primary null: R² under randomised B_obs (999 permutations)", eval=!is.null(null_sum) && length(null_sum$figures) >= 1}
knitr::include_graphics(null_sum$figures[1])
```

```{r fig-null-robust, fig.cap="Robustness null: R² under B shuffle across hauls", eval=!is.null(null_sum) && length(null_sum$figures) >= 2}
knitr::include_graphics(null_sum$figures[2])
```

---

## 5. Summary checklist

```{r checklist}
checks <- tibble(
  Check = c(
    "DATRAS state variables built (build_datras_state_variables.R)",
    "EEoS predictions built (build_eeos_predictions.R)",
    "ln(E) reference fitted (run_h1_lne_reference.R)",
    "EEoS vs ln(E) compared",
    "S == n_species_with_lw (DATRAS consistency)",
    "B-null scheme chosen from log(B) shape",
    "Null permutations complete (999)",
    "Primary residual = log(B_obs) - log(B_pred)"
  ),
  Status = c(
    if (file.exists(file.path(project_root, "outputs", "datras_haul_state.rds"))) "PASS" else "PENDING",
    if (file.exists(path_preds)) "PASS" else "PENDING",
    if (file.exists(path_compare)) "PASS" else "PENDING",
    if (file.exists(path_compare)) "PASS" else "PENDING",
    if (all(haul$n_species_with_lw == haul$S, na.rm = TRUE)) "PASS" else "FAIL",
    "PASS",
    if (!is.null(null_sum)) "PASS" else "PENDING",
    "PASS"
  )
)
kable(checks) |> kable_styling(full_width = FALSE)
```

---

## 6. Interpretation (supervisor discussion)

```{r interpretation, results='asis'}
if (!is.null(compare) && !is.null(null_sum)) {
  r2_eeos <- compare$comparison$log_r2[compare$comparison$model_id == "eeos_biomass"]
  r2_lne <- compare$comparison$log_r2[compare$comparison$model_id == "ln_e_ols"]
  ratio <- compare$ss_ratio_eeos_over_lne

  cat("1. **Absolute EEoS performance:** log-scale R² =", sprintf("%.2f", r2_eeos),
      "with median |residual| =", sprintf("%.2f", null_sum$median_abs_residual_obs),
      "and ~4× median underprediction (B_pred/B_obs).\n\n")
  cat("2. **vs ln(E) benchmark (Harte et al. 2022):** ln(E) alone achieves R² =",
      sprintf("%.2f", r2_lne), "vs EEoS R² =", sprintf("%.2f", r2_eeos),
      ". Unexplained SS is", sprintf("%.1f×", ratio),
      "higher for EEoS. Harte criterion **not met** — S and N do not improve on metabolic rate alone at haul scale.\n\n")
  cat("3. **vs B-null:** Observed EEoS pairing beats random biomass labels (primary uniform null p ≈",
      sprintf("%.3f", null_sum$primary$p_value_one_sided), ") but both models have poor absolute R².\n")
}
```
